---
title: "Scototaxis_walk-through"
author: "Allison Davis"
date: "May 6, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data collection

You first need to call you data from the internet in order to run stats in Rstudio. Your data should be saved as a csv file, and uploaded to your github repository (github project). On github, click on your project, click on the csv file, then hit "raw" --you will need to copy the html of the raw file to use in Rstudio.

To pull your data into Rstudio, you will need to use the following functions/packages: library(), curl package, curl(), read.csv()

  * If you haven't done so already, install the curl package. You can install a package in two ways:
      1) use `install.packages()` with the name of the package in the brackets
      
      2) go to your packages tab in the lower right corner (Files Plots **Packages** Help Viewer), hit install, and type in the name of the package before hitting the install button again.
  
  * Stuck? See the "Data collection" section in the Intro to R html.


```{r}
#load in the curl package


#pull the data from the internet using your html and assign a name


#force your data to be read as a csv


```

Now we want to make sure your data was pulled correctly, and we also want to create two separate data frames for each of the species. This way we can do stats on three data frames: raw data (original file we pulled from the internet), sailfin (all the info but just for P. latipinna), and amazon (all the info but just for P. formosa).

For this step, you will need: head(), data frame selection (aka data frame[data frame$name .....])

  * Stuck? See the "Data collection" section in the Intro to R html (part about creating a data frame with only certain elements)
  
  * Make sure you assign a new name for your new data frames! If you keep the name the same for all three data frames, R will overwrite (aka erase) your data!
  
```{r}
#check the head of your raw dataframe to make sure it was read in correctly
#NOTE: whenever you read in a csv, R automatically adds X.U.FEFF to the first column. This is normal!


#create a dataframe for sailfin by selecting data from your raw dataframe with P. latipinna as the species.


#create a dataframe for amazons by selecting data from your raw dataframe with P. formosa as the species.


#check the head of your new sailfin and amazon dataframes to make sure it selected correctly.



```


## Data analyses

Now we get to the fun part -actually analyzing our data! While most people want to jump right in to the most complicated stat models, it's actually better practice to get an idea of what your data look like. Getting summary statistics (mean, median, mode etc) and visualizing your data give you an idea of how your data relate to each other. This will make the conclusions from the more complicated stats easier to understand.

So let's begin with summary statistics. For this you will need: rstatix package, get_summary_stats()

  * Make sure to install the rstatix package. See the beginning of this Rmd on how to install.
  
  * The summary stats you want are: mean, stadard deviation, median, and interquartile range.
  
  * For more details on how to use get_summary_stats() see: https://www.datanovia.com/en/lessons/wilcoxon-test-in-r/#summary-statistics-2
  
      + ex: get_summary_stats(weight, type ="median", data=mice)
          + the %>% used on the website requires another package, but is just specifying the data frame to use; instead use data=name in fuction.


```{r}
#load in the rstatix package 


#get summary statistics for all data


#get summary statistics for the sailfin data  #NOTE: want to get an idea of how each species looks, so we'll separate them.


#get summary statistics for the amazon data



```


While numbers are great, a picture is often easier to understand. So let's graph out our data to get a better idea of what's going on. Box plots are a good graph to use, since it will show us the mean and range of the data. 

At the moment, let's focus on duration data. Our main goal is to see if the fish have a preference for a particular color.

For this you will use: boxplot()

  * For more info on boxplot() see: https://www.statmethods.net/graphs/boxplot.html
  
  * We want our boxplots to display duration in one color by tank AND species. When we want to display our y value (duration) by two x values (tank and species), we denote this as y~x1*x2
  
  * Remember to add x and y lables! 


```{r}
#create a box plot showing duration in white by tank and species


#create a box plot showing duration in black by tank and species



```


Now that you've done basic statistics, how does your data look? Write about the trends you see below.

<font color="deeporchid4"> ***First look at data:*** </font>









Ok, now we dive even deeper. Let's review the questions we want to answer with our more complicated stats:

1) Are there differences between tank A and tank B? We want to make sure there wasn't a tank bias, meaning they all acted weird in tank B over tank A. We also want to make sure that there isn't a trial bias, meaning 






